Hypercoercions and a Framework for Equivalence of Cast Calculi

### 2019-10-23 paper.tex
finished compact representation paragraph

This paper presents a new cast representation, named
\emph{hypercoercions}, that is good for both mechanized metatheory and
good for implementation. The composition operator for hypercoercions
is defined by structural recursion and hypercoercions are suggestive
of a bit-level representation that minimizes the need for pointers and
fits all first-order casts into 64 bits. We present two flavors of
hypercoercions to support the two blame tracking strategies from the
literature: D and UD~\citep{siek2009exploring}. We are interested in
the D blame tracking strategy because it comes with a more
straightforward notion safe cast compared to
UD~\citep{siek2009exploring}, which is why D was chosen Grift
compiler. We are also intersted in UD because it plays a prominent
role in the gradual typing literature \citep{wadler2009well}. The
semantics of casts can also be lazy or
eager~\citep{siek2009exploring}. In this paper we focus on lazy cast
strategies because we suspect that they are more efficient than eager
strategies and because \citet{new2019gradual} show that eager
strategies are incompatible with $\eta$-equivalence of functions.

@@

% the following are temporary -JS
\clearpage
\pagebreak

@@

\begin{figure}
  Syntax
  \[
  \begin{array}{lclr}
  \stxrule{I}{injectable types (\lazyD)}{P}
  \stxrule{I}{injectable types (\lazyUD)}{
    \POOunit \mid
    \POOfun{\TOOdyn}{\TOOdyn} \mid
    \POOprod{\TOOdyn}{\TOOdyn} \mid
    \POOsum{\TOOdyn}{\TOOdyn}
  }
  \stxrule{c}{hypercoercions}{
    \hyperCoercionI \mid{}
    \hyperCoercionC{h}{m}{t}
  }
  \stxrule{h}{heads}{
    \epsilon \mid{}
    ?^l
  }
  \stxrule{m}{middles}{
    \POOunit \mid
    \POOfun{c_1}{c_2} \mid
    \POOprod{c_1}{c_2} \mid
    \POOsum{c_1}{c_2}
  }
  \stxrule{t}{tails}{
    \epsilon \mid{}
    ! \mid{}
    \bot^l
  }
  \end{array}
  \]

@@

\begin{proposition}[\lazyD\ hypercoercions form a monoid]
  \label{thm:hc-monoid}
  For all $c : T_1 \Longrightarrow T_2$,
  $c_1 : T_1 \Longrightarrow T_2$,
  $c_2 : T_2 \Longrightarrow T_3$, and
  $c_3 : T_3 \Longrightarrow T_4$,
  \begin{enumerate}
    \item $seq(id(T_1),c) = c$,
    \item $seq(c,id(T_2)) = c$, and
    \item $seq(seq(c_1, c_2), c_3) = seq(c_1, seq(c_2, c_3))$.
  \end{enumerate}
\end{proposition}
\begin{proof}
  Part (1) and (2) are by induction on $c$. Part (3) is by induction
  on $c_2$ and case analysis on $c_1$ then $c_3$.
\end{proof}

\begin{proposition}[Identity Casts]
  For all $T$ and $l$, $cast(T,l,T) = id(T) $
\end{proposition}
\begin{proof}
  By induction on T.
\end{proof}

\subsection{\lazyUD{} hypercoercion [Jeremy]}
\label{sec:lud-hc}

@@

\begin{proposition}[\lazyUD\ hypercoercions form a monoid]
  For all $c : T_1 \Longrightarrow T_2$,
  $c_1 : T_1 \Longrightarrow T_2$,
  $c_2 : T_2 \Longrightarrow T_3$, and
  $c_3 : T_3 \Longrightarrow T_4$,
  \begin{enumerate}
    \item $seq(id(T_1),c) = c$,
    \item $seq(c,id(T_2)) = c$, and
    \item $seq(seq(c_1, c_2), c_3) = seq(c_1, seq(c_2, c_3))$.
  \end{enumerate}
\end{proposition}

@@

\todo[inline]{I need to add $cast(T,l,T) = id(T)$. -JS}

\subsection{Compact Memory Representation for Hypercoercions}

\todo[inline]{UNDER CONSTRUCTION (a couple paragraphs) -Jeremy}

% the following are temporary -JS
\clearpage
\pagebreak

>>>>

This paper presents a new cast representation, named
\emph{hypercoercions}, that is good for both mechanized metatheory and
good for implementation. The composition operator for hypercoercions
is defined by structural recursion and hypercoercions are suggestive
of a bit-level representation that minimizes the need for pointers and
fits all first-order casts into 64 bits.
%
We present two flavors of hypercoercions to support the two blame
tracking strategies from the literature: D and
UD~\citep{siek2009exploring}. We are interested in the D blame
tracking strategy because it comes with a more straightforward notion
safe cast compared to UD~\citep{siek2009exploring}, which is why D was
chosen Grift compiler. We are also intersted in UD because it plays a
prominent role in the gradual typing literature
\citep{wadler2009well}.  The UD hypercoercions were inspired by the
supercoercions of \citet{garcia2013calculating} (hence the name) and
the D hypercoercions were inspired by the normal forms of
\citet{siek2012interpretations}.
%
The semantics of casts can be lazy or
eager~\citep{siek2009exploring}. In this paper we focus on lazy cast
strategies because we suspect that they are more efficient than eager
strategies and because \citet{new2019gradual} show that eager
strategies are incompatible with $\eta$-equivalence of functions.

@@

@@

\begin{figure}
  Syntax of Hypercoercions
  \[
  \begin{array}{lclr}
  \stxrule{I}{injectable types (\lazyD)}{P}
  \stxrule{I}{injectable types (\lazyUD)}{
    \POOunit \mid
    \POOfun{\TOOdyn}{\TOOdyn} \mid
    \POOprod{\TOOdyn}{\TOOdyn} \mid
    \POOsum{\TOOdyn}{\TOOdyn}
  }
  \stxrule{c}{hypercoercions}{
    \hyperCoercionI \mid{}
    \hyperCoercionC{h}{m}{t}
  }
  \stxrule{h}{heads}{
    \epsilon \mid{}
    ?^l
  }
  \stxrule{m}{middles}{
    \POOunit \mid
    \POOfun{c_1}{c_2} \mid
    \POOprod{c_1}{c_2} \mid
    \POOsum{c_1}{c_2}
  }
  \stxrule{t}{tails}{
    \epsilon \mid{}
    ! \mid{}
    \bot^l
  }
  \end{array}
  \]

@@

\begin{proposition}[\lazyD{} hypercoercions form a monoid]
  \label{thm:hc-monoid}
  For all $c : T_1 \Longrightarrow T_2$,
  $c_1 : T_1 \Longrightarrow T_2$,
  $c_2 : T_2 \Longrightarrow T_3$, and
  $c_3 : T_3 \Longrightarrow T_4$,
  \begin{enumerate}
    \item $seq(id(T_1),c) = c$,
    \item $seq(c,id(T_2)) = c$, and
    \item $seq(seq(c_1, c_2), c_3) = seq(c_1, seq(c_2, c_3))$.
  \end{enumerate}
\end{proposition}
\begin{proof}
  Part (1) and (2) are by induction on $c$. Part (3) is by induction
  on $c_2$ and case analysis on $c_1$ then $c_3$.
\end{proof}

\begin{proposition}[\lazyD{} Identity Casts]
  For all $T$ and $l$, $cast(T,l,T) = id(T) $
\end{proposition}
\begin{proof}
  By induction on T.
\end{proof}

\subsection{\lazyUD{} hypercoercions}
\label{sec:lud-hc}

@@

\begin{proposition}[\lazyUD{} hypercoercions form a monoid]
  For all $c : T_1 \Longrightarrow T_2$,
  $c_1 : T_1 \Longrightarrow T_2$,
  $c_2 : T_2 \Longrightarrow T_3$, and
  $c_3 : T_3 \Longrightarrow T_4$,
  \begin{enumerate}
    \item $seq(id(T_1),c) = c$,
    \item $seq(c,id(T_2)) = c$, and
    \item $seq(seq(c_1, c_2), c_3) = seq(c_1, seq(c_2, c_3))$.
  \end{enumerate}
\end{proposition}
\begin{proof}
  See the Agda proof at the following URL:

  \url{https://github.com/jsiek/gradual-typing-in-agda/blob/master/HyperCoercions.agda}

\end{proof}

@@

\begin{proposition}[\lazyUD{} Identity Casts]
  For all $T$ and $l$, $cast(T,l,T) = id(T) $
\end{proposition}
\begin{proof}
  By induction on T.
\end{proof}

\subsection{Compact Representation of Hypercoercions}

Hypercoercions enable a bit-level representation that is particularly
compact for the most common-case coercions. The most frequent
coercions are injections and projections to base types (here
$\POOunit$ is the only base type, but in a real language the base
types would include integers, Booleans, etc.) and identity coercions.
Values of base types are often stored in CPU registers, so it would be
nice for coercions on base types to also fit in registers, that is, in
64 bits, so that applying a coercion to a value would not require
access to main memory, which is an order of magnitude slower than
accessing registers on a modern CPU.
%
Coercions involving non-base types, such as functions types,
may be arbitrarily large, so in those cases the hypercoercion
representation can be a pointer to a heap-allocated structure.  Here
is a sketch for the bit-level representation for hypercoercions.

\begin{itemize}
\item 1 bit to differentiate between $\hyperCoercionI$
  and three-part coercions.

\item 1 bit to differentiate between middle coercions of base type
  versus non-base type.

\item If the middle is of non-base type, then 61 bits represent a
  pointer to a heap-allocated structure. Heap allocated structures are
  usually 8-byte aligned, so there are 3 unused bits in a pointer.

\item If the middle is of base type, then the remaining 62 bits are
  used to represent the head (21 bits), middle (22 bits), and tail (19
  bits).
  \begin{itemize}
  \item The head requires 1 bit to differentiate between $\epsilon$
    and $?^l$ and then 24 bits could be used for the label $l$.
  \item The middle would use 11 bits to differentiate all the base types.
  \item The tail requires 2 bits to differentiate between $\epsilon$,
    $!$, and $\bot^l$, and could use 24 bits for the label $l$.
  \end{itemize}
  In the event that a program requires too many blame labels, then one
  can fallback to using the heap-allocated structure for more
  coercions.
\end{itemize}
